---
title: "Subsetting in R Solutions"
author: "Julianne G. Clina"
institute: "University of Kansas Medical Center"
date: 2026-08-26
format:
  html:
    toc: true
    toc-depth: 2
    code-block-bg: true
    theme: cosmo
execute:
  echo: true
  eval: false
---

## Learning Objectives

By the end of this workshop, you should be able to:

- Explain the difference between `[`, `[[`, and `$`, and know when to use each
- Subset a data frame by rows and columns using `data[rows, columns]`
- Explain the difference between **simplifying** and **preserving** subsetting
- Build logical conditions with `==`, `>=`, `<`, `&`, `|`, and `!` to select rows
- Combine multiple conditions (including with parentheses) to answer a realistic research question
- Translate a plain-English data request into R subsetting code

The core question to ask yourself throughout: **what information do I want, and where is it stored?**

```{r}
##Creating Example Dataset for practice!
library(tidyverse)


kids <- tibble(
  id = sprintf("K%02d", 1:15),
  age = c(
    8,
    9,
    10,
    8,
    11,
    12,
    9,
    10,
    11,
    8,
    12,
    9,
    10,
    11,
    12
  ),
  sex = c(
    "F",
    "M",
    "F",
    "M",
    "F",
    "M",
    "F",
    "F",
    "M",
    "M",
    "F",
    "M",
    "F",
    "M",
    "F"
  ),
  bmi_z = c(
    -0.2,
    0.8,
    1.4,
    -0.5,
    2.1,
    0.3,
    1.1,
    -0.8,
    0.6,
    1.8,
    0.2,
    1.3,
    -0.1,
    2.4,
    0.9
  ),
  mvpa_min = c(
    62,
    41,
    28,
    75,
    35,
    54,
    92,
    67,
    46,
    22,
    81,
    38,
    70,
    31,
    55
  ),
  sedentary_min = c(
    510,
    560,
    610,
    480,
    590,
    525,
    450,
    500,
    545,
    625,
    470,
    580,
    495,
    600,
    530
  ),
  steps = c(
    9200,
    7100,
    5400,
    10400,
    6300,
    8100,
    12000,
    9800,
    7600,
    4900,
    11000,
    6800,
    10100,
    5700,
    8300
  ),
  sbp = c(
    102,
    108,
    115,
    98,
    121,
    110,
    105,
    100,
    112,
    118,
    107,
    114,
    103,
    124,
    109
  ),
  dbp = c(
    64,
    68,
    72,
    61,
    76,
    70,
    66,
    62,
    69,
    74,
    67,
    71,
    65,
    78,
    68
  ),
  resting_hr = c(
    82,
    78,
    88,
    76,
    91,
    80,
    75,
    79,
    84,
    94,
    77,
    86,
    74,
    90,
    81
  ),
  vo2peak = c(
    42.1,
    38.4,
    34.2,
    45.6,
    31.8,
    40.2,
    47.1,
    43.5,
    37.9,
    29.4,
    44.8,
    35.6,
    46.2,
    32.5,
    39.7
  )
)

kids$parent_info <- list(
  list(mother = "Sarah", father = "James", contact = "email"),
  list(mother = "Maria", father = "Carlos", contact = "phone"),
  list(mother = "Jennifer", father = "David", contact = "email"),
  list(mother = "Amanda", father = "Brian", contact = "phone"),
  list(mother = "Rachel", father = "Michael", contact = "email"),
  list(mother = "Lisa", father = "Robert", contact = "email"),
  list(mother = "Emily", father = "Daniel", contact = "phone"),
  list(mother = "Katie", father = "Andrew", contact = "email"),
  list(mother = "Nicole", father = "Matthew", contact = "phone"),
  list(mother = "Jessica", father = "Christopher", contact = "email"),
  list(mother = "Laura", father = "Steven", contact = "phone"),
  list(mother = "Michelle", father = "Kevin", contact = "email"),
  list(mother = "Elizabeth", father = "Mark", contact = "email"),
  list(mother = "Heather", father = "Jason", contact = "phone"),
  list(mother = "Rebecca", father = "Thomas", contact = "email")
)

kids$mvpa_min[c(5, 12)] <- NA
kids$sbp[9] <- NA
kids$bmi_z[13] <- NA

kids
```

---

## Activity 1: `[`, `[[`, and `$`

### Why three operators?

R gives you three tools for pulling things out of an object, and they are **not interchangeable**:

| Operator | What it returns | Typical use |
|---|---|---|
| `[`  | A **subset** of the same type/structure (can hold 0, 1, or many elements) | "Give me a smaller version of this object" |
| `[[` | A **single element**, unwrapped from its container | "Give me the contents of this one slot" |
| `$`  | A single **named** element (shorthand for `[[` with a name) | "Give me the column/element called ___" |

A helpful mental picture from *Advanced R* (Wickham): if a list is a **train**, and each element is a **train car**, then:

- `x[4:6]` gives you a smaller train made of cars 4 through 6 (still a train — a list)
- `x[[5]]` opens car 5 and hands you what's *inside* it (not a train anymore)

This is why `[` on a list always returns a list (even a list of length 1), while `[[` and `$` reach inside and hand you the actual object stored there.

Our `parent_info` column is a **list-column** — each element of `kids$parent_info` is itself a list. It's a good place to see `[` vs `[[` vs `$` in action.

### Step 1: Look at the list-column

```{r}
kids$parent_info
```

### Step 2: Extract one child's parent information

Suppose we want the parent information for child 3.

```{r}
# [ ] extracts the third element as a length-1 list
kids$parent_info[3]

# [[ ]] extracts the third element itself (a named list, not wrapped in a list)
kids$parent_info[[3]]

class(kids$parent_info[3])
class(kids$parent_info[[3]])
```

**What's different?** `[` keeps the "train car" wrapper — you get a list containing one list. `[[` unwraps it and gives you the list itself. This is the **simplify vs. preserve** distinction: `[[` simplifies (drops a layer of structure), `[` preserves it.

### Step 3: Extract information within the list

Now that we have the parent information for child 3, we want to extract just the mother's name. Because `parent_info[[3]]` is a *named* list, we can use `$` to grab a named piece of it.

```{r}
kids$parent_info[[3]]$mother
```

The key idea: these operators **chain together** to walk through nested objects — first `[[` to get into the list-column, then `$` to get the named element inside.

---

## Activity 2: Selecting Rows and Columns

Data frames behave like a cross between a list (each column is an element) and a matrix (rows and columns). When subsetting with two indices, the pattern is:

```r
data[rows, columns]
```

- The first position (before the comma) says **which rows**
- The second position (after the comma) says **which columns**
- Leaving a position blank means "give me all of them"

```{r}
# All columns for child 5
kids[5, ]

# The mvpa_min column, for all children
kids[, "mvpa_min"]
```

### Your turn

Imagine you are preparing a small table for a research meeting. You want information for children 2, 5, 8, and 12 — but only the columns `id`, `age`, `sex`, `mvpa_min`, and `vo2peak`.

```{r}
# Select children 2, 5, 8, and 12
# and keep only id, age, sex, mvpa_min, and vo2peak

kids[
  c(2, 5, 8, 12),
  c("id", "age", "sex", "mvpa_min", "vo2peak")
]
```

### Simplify vs. preserve: one more thing to notice

Compare these two approaches to grabbing a single column:

```{r}
kids[, "mvpa_min"] # matrix-style: simplifies
kids["mvpa_min"] # list-style: preserves
```

They select the same values, but return **different types of objects**:

```{r}
class(kids[, "mvpa_min"])
class(kids["mvpa_min"])
```

`kids[, "mvpa_min"]` simplifies down to a plain vector, because you asked for it "matrix style" (row, column). `kids["mvpa_min"]` treats `kids` like a list and hands you back a smaller data frame with one column, preserving the original structure. This distinction is a common source of bugs — a function that works fine when you select multiple columns can break unexpectedly when only one column is selected, because the object type silently changes. (Tibbles were designed to avoid this: `kids[, "mvpa_min"]` on a tibble still returns a tibble, not a bare vector — try it and compare to a base `data.frame`.)

---

## Activity 3: Logical Subsetting

So far we've selected rows by their *position* (row 5, rows 2/5/8/12). But in research data, we usually want to select rows by whether they meet a **condition** — "give me children who did X" rather than "give me row 7."

This is where logical subsetting comes in. A logical condition like:

```{r}
kids$mvpa_min >= 60
```

produces a vector of `TRUE`, `FALSE`, and possibly `NA` — one value per row. When you put that logical vector inside `[`, R keeps only the rows where the value is `TRUE`, and returns `NA` for rows where the condition itself was `NA` (this is different from using `which()`, which silently drops `NA`s instead — worth knowing when you're not sure why your row counts don't match).

### Your turn

Find the children who get at least 60 minutes of MVPA. Return only `id`, `age`, `mvpa_min`, and `vo2peak`.

```{r}
# Select children with at least 60 minutes of MVPA
# Keep only id, age, mvpa_min, and vo2peak

kids[
  kids$mvpa_min >= 60,
  c("id", "age", "mvpa_min", "vo2peak")
]
```

### Try another one

Now find just the **IDs** of children who get less than 30 minutes of MVPA.

```{r}
# Select the IDs of children with MVPA < 30

kids$id[kids$mvpa_min < 30]
```

Notice this time we're subsetting a **vector** (`kids$id`), not the whole data frame — same logical-subsetting idea, just applied outside the `data[rows, columns]` pattern.

---

## Activity 4: Combining Conditions

Multiple logical conditions can be combined with:

| Operator | Meaning |
|---|---|
| `&` | AND (both conditions must be TRUE) |
| `|` | OR (at least one condition must be TRUE) |
| `!` | NOT (flips TRUE/FALSE) |

Always use the vectorized `&` and `|` here (not `&&` / `||`, which only compare a single value each and are meant for `if` statements).

```{r}
kids$mvpa_min >= 60 &
  kids$bmi_z < 1
```

### Your turn

Identify children who have **either** low physical activity (MVPA < 30 minutes) **or** high BMI z-score (≥ 2). Return `id`, `age`, `bmi_z`, `mvpa_min`.

```{r}
# Select children with either:
#   MVPA < 30
#   OR BMI z-score >= 2
#
# Keep id, age, bmi_z, and mvpa_min

kids[
  kids$mvpa_min < 30 |
    kids$bmi_z >= 2,
  c("id", "age", "bmi_z", "mvpa_min")
]
```

### Add another condition

Now restrict to **girls** who meet either of those criteria:

> Female AND (low MVPA OR high BMI z-score)

```{r}
# Select girls who either:
#   have MVPA < 30
#   OR have BMI z-score >= 2
#
# Keep id, age, sex, bmi_z, and mvpa_min

kids[
  kids$sex == "F" &
    (kids$mvpa_min < 30 | kids$bmi_z >= 2),
  c("id", "age", "sex", "bmi_z", "mvpa_min")
]
```

**Why the parentheses matter:** without them, R would evaluate `&` before `|` in a way that changes the meaning of the statement (operator precedence). The parentheses force the OR condition to be evaluated as a single unit before it's combined with the AND. When in doubt, use parentheses to make your intent explicit — even when they're not strictly required, they make code easier for you (and others) to read later.

---

## Activity 5: Putting It All Together

Your PI sends you this request:

> "Can you give me the ID, age, sex, MVPA, BMI z-score, and systolic blood pressure for girls who either get less than 30 minutes of MVPA or have a BMI z-score of 2 or greater? Please exclude anyone with missing systolic blood pressure."

Before writing any code, translate the request into pieces.

**Which rows do we want?**

- Girls
- AND either: MVPA < 30, OR BMI z-score ≥ 2
- AND systolic BP is not missing

**Which columns do we want?**

- id, age, sex, mvpa_min, bmi_z, sbp

### Your turn

```{r}
# Your code here

kids[
  kids$sex == "F" &
    (kids$mvpa_min < 30 | kids$bmi_z >= 2) &
    !is.na(kids$sbp),
  c(
    "id",
    "age",
    "sex",
    "mvpa_min",
    "bmi_z",
    "sbp"
  )
]
```

Notice `!is.na(kids$sbp)` — `is.na()` returns `TRUE` where a value is missing, so `!is.na()` flips that to `TRUE` where a value is **not** missing (i.e., where you want to keep the row).

---
